---
title: "Statistical Inference Course Project - Part 2"
author: "Aaron Cole"
date: "8 February 2016"
output:
  pdf_document:
    number_sections: yes
    toc: yes
  html_document:
    toc: yes
---

# Synopsis #
The following text is quoted from the assignment information page on Coursera.

> * Load the ToothGrowth data and perform some basic exploratory data analyses.
> * Provide a basic summary of the data.
> * Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering).
> * State your conclusions and the assumptions needed for your conclusions.

Evaluation criteria as follows.

> * Did you perform an exploratory data analysis of at least a single plot or table highlighting basic features of the data?
> * Did the student perform some relevant confidence intervals and/or tests?
> * Were the results of the tests and/or intervals interpreted in the context of the problem correctly?
> * Did the student describe the assumptions needed for their conclusions?

# Exploratory analysis and basic summary #

## Load libraries ##
```{r echo=TRUE}
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
```

## Load tooth growth data and configure groupings ##
```{r echo=TRUE}
data(ToothGrowth)
tg <- tbl_df(ToothGrowth)
tg <- group_by(tg, supp, dose)
```

## Summary and explanation of the data ##
```{r echo=TRUE}
summary(tg)
```

The data is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid). The data frame has 60 observations on 3 variables.

Variable | Class | Comments
---------|-------|---------
len	| numeric | Tooth length.
supp | factor | Supplement type (VC or OJ).
dose | numeric | Dose in milligrams.

## Exploratory analysis and basic summary ##
```{r echo=TRUE}
ggplot(ToothGrowth, aes(x=factor(dose),
                    y=len,
                    fill=factor(dose))) +
    geom_boxplot(notch=F) + 
    facet_grid(.~supp) +
    ggtitle("Tooth length split by dose for each delivery method") +
    xlab("Dose") +
    ylab("Length") +
    scale_fill_discrete(name="Dose")

# Text summary of the means by supp and dose.
summarise_each(tg, funs(mean))

# Text summary of the total growth (len) by supp.
summarise_each(select(group_by(tg, supp), supp, len), funs(sum))
```

* The plot and text summary appear to indicate that higher dose levels result in higher growth length.
* Supplement type "OJ" would appear to result in higher average growth at the lower dose levels of 0.5 and 1.0.
* Supplement type "OJ" also has higher growth across all observations for all dose levels, when compared with "VC".
* The average growth is similiar for both supplement types at a dose level of 2.0.

# Using confidence intervals to compare tooth growth by supp and dose #

## Assumptions ##

* Our data is roughly normal.
* The variance between groups is unequal.

## Testing the supplements for all dose levels##
```{r echo=TRUE}
t.test(len ~ supp, 
       data = filter(tg), 
       paired = FALSE, 
       var.equal = TRUE, 
       alternative = "two.sided")
```

* P-value is above 5%.
* Confidence interval contains 0.

## Testing the supplements at a dose level of 0.5 ##
```{r echo=TRUE}
t.test(len ~ supp, 
       data = filter(tg, dose == 0.5), 
       paired = FALSE, 
       var.equal = TRUE, 
       alternative = "two.sided")
```

* P-value is below 5%.
* Confidence interval does not contain 0.
* Indicates with 95% confidence that supplement type OJ results in higher growth rates at a dose of 0.5.

## Testing the supplements at a dose level of 1.0 ##
```{r echo=TRUE}
t.test(len ~ supp, 
       data = filter(tg, dose == 1.0), 
       paired = FALSE, 
       var.equal = TRUE, 
       alternative = "two.sided")
```

* P-value is below 5%.
* Confidence interval does not contain 0.
* Indicates with 95% confidence that supplement type OJ results in higher growth rates at a dose of 1.0.

## Testing the supplements at a dose level of 2.0 ##
```{r echo=TRUE}
t.test(len ~ supp, 
       data = filter(tg, dose == 2.0), 
       paired = FALSE, 
       var.equal = TRUE, 
       alternative = "two.sided")
```

* P-value is above 5%.
* Confidence interval contains 0.

## Conculsions ##

* Our t-tests support a difference between the supplements only for dose levels of 0.5 and 1.0, with a 95% confidence.
* At a dose level of 0.5 and 1.0 supplement type OJ appears to promote higher growth.
* Our t-tests do not support any difference between the supplements at a dose level of 2.0, or across the total sample for all dose levels.